Time for a rethink: disparity-through-time

### 2017-12-08 manuscript/time-slice.tex
grant up
Edits to end of time slice methods

To address these issues, we propose a ``time-slicing'' approach that takes advantage of the wealth of paleontological datasets which now have associated phylogenies. 
Time slicing uses a phylogenetic tree and considers subsets of taxa at specific equidistant points in time, as opposed to considering subsets of taxa between two points in time.
This results in even-sampling across time and permits us to define the underlying model of character evolution (punctuated or gradual).  
Time-slicing also includes any element present in the phylogeny (branches, nodes and tips) at the time slice in question as part of the disparity calculation.
This allows us to measure disparity at time points where there are no terminal taxa, and increases the sample size at each time point, making downstream analyses of the drivers of disparity much more feasible.

Here we present our time slicing methods using four datasets taken from the literature.
We calculate disparity-through-time for each dataset using a range of time binning and time slicing methods, and then compare these approaches with respect to the relative disparities calculated, but also with how the different approaches influence biological conclusions. We find that XXX

\section{Materials and Methods}
\subsection{Overview}
\label{overview-section}
% I'm unsure if this section is needed or not. The methods are long so it might help?
%TG: agree though we might just do a simple figure like fig1 in http://onlinelibrary.wiley.com/doi/10.1111/bij.12746/pdf (but with way less stuff!). I can get working on that.
To test the different time subsampling methods, we followed the protocol below (Figure). 
All the code needed to reproduce these analyses (along with detailed instructions) is provided on GitHub (https://github.com/nhcooper123/time-slice). % NC: might move this to the NHM github account

%\begin{enumerate}
 % \item Collating example datasets. Disparity-through-time analysis requires a matrix containing morphological characters for each taxon, and information on the first and last occurrence dates of each taxon. Time slicing methods also require a phylogeny.
  %\item Processing example datasets. Various transformations of the datasets are required before the analyses can take place. We need to estimate ancestral character states at the nodes of the phylogeny for the time slicing methods, build cladisto-spaces via distance matrices and ordination, and remove taxa from the phylogeny that are missing from the matrix and vice versa.   
  %\item Time subsampling
  %\item Estimate disparity for each time subsample
  %\item Compare different methods
%\end{enumerate}

\subsection{Example datasets}
\label{datasets}
To test the different time binning/slicing methods we selected four datasets: a mammal dataset from \cite{beckancient2014}, two theropod datasets from \cite{brusatte2014gradual} and \cite{bapst2016topology}, and a crinoid dataset from \cite{wright2017bayesian}.
Table \ref{datasets} and appendix ??? provide more details. % add appendix details
%Note that although disparity-through-time analyses are possible without a phylogeny, we only use examples with phylogenies here so that we can compare results of time slicing methods, which do require a phylogeny. 
Each dataset consists of first and last occurrence dates for all taxa, a matrix of morphological characters in NEXUS format, and a time-scaled phylogeny. 
These datasets are freely available with their accompanying papers (Table \ref{datasets}), but for reproducibility purposes we also provide the data we used on the NHM Data Portal (CITE).

  % \begin{table}[!htbp]      
  %   \caption[Datasets.]
  %   {Dataset captions}
    \input{tables/table_datasets.tex} 
    \label{datasets}  
  % \end{table}

\subsection{Preparing the data for disparity-through-time analysis}
\label{ace}

\paragraph{Estimating ancestral character states.}
For each dataset we used the re-rooting method \citep{Yang01121995,Garland2000} to get Maximum Likelihood estimates of the ancestral states for each character at every node in the phylogeny, using the \texttt{rerootingMethod} function from the \texttt{R} package \texttt{phytools} version 0.6-44 \citep{phytools,R}.
Where there were missing character data for a taxon we followed the method of \cite{Claddis} and treated missing data as any possible observed state for each character.
For example, if a character had two observed states ($0$ and $1$) across all taxa, we attributed the multi-state ``$0$\&$1$" value to the taxon with missing data, representing an equal probability of being either $0$ or $1$.
This allows the ancestral node of a taxon with missing data to be estimated with no assumptions other than that the taxon has one of the observed character states.
To prevent poor ancestral state reconstructions from biasing our results, especially when a lot of error is associated with the reconstruction, we only included ancestral state reconstructions with a scaled Likelihood $\geq$ $0.95$.
Ancestral state reconstructions with scaled Likelihoods below this threshold were recoded as missing data (``?'').

@@

\paragraph{Constructing distance matrices.}
To estimate the cladisto-spaces for each of our datasets we first constructed pairwise distance matrices of length $k$, where $k$ is the total number of tips and nodes in the dataset.
We calculated the $k$$\times$$k$ distances using the Gower distance \citep{Gower71}, i.e. the Euclidean distance between two taxa divided by the number of shared characters. 
This allows us to correct for distances between two taxa that share many characters and could be closer to each other than to taxa with fewer characters in common (i.e. because some pairs of taxa share more characters in common than others, they are more likely to be similar).
For cladistic matrices, using this corrected distance is preferable to the raw Euclidean distance because of its ability to deal with discrete or/and ordinated characters as well as with missing data \citep{anderson2012using}.
However, the Gower distance cannot calculate distances when taxa have no overlapping data.
Therefore, we used the \texttt{TrimMorphDistMatrix} function from the \texttt{Claddis} \texttt{R} package \citep{Claddis} to remove pairs of taxa with no cladistic characters in common.
This led to us removing nine taxa from the \cite{bapst2016topology} dataset, and 19 from the \cite{brusatte2014gradual} dataset, but none from the other two datasets (see appendix for details of which species). % add appendix

@@

\subsection{Disparity-through-time analyses}
\label{disparity_calc}

\paragraph{Calculating disparity.}
Disparity can be calculated in many different ways \citep[e.g.][]{Wills1994,Ciampaglio2004,thorneresetting2011,hopkinsdecoupling2013,huang2015origins}, however most studies estimate disparity using four metrics: the sum and products of ranges and variances, each of which gives a slightly different estimate of how the data fits within the cladisto-space \citep{Foote01071994,Wills1994,brusatte50,Brusatte12092008,cisneros2010,thorneresetting2011,prentice2011,brusattedinosaur2012,toljagictriassic-jurassic2013,ruta2013,bentonmodels2014,bensonfaunal2014}.
However, these metrics have limitations. 
First, the range metrics are affected by the uneven sampling of the fossil record \citep{Butler2012}.
Second, because we include all $n$ dimensions in the analysis (see above), the products of ranges and variances will tend towards zero since the scores of the last dimension are usually really close to zero themselves. 
We therefore use the sum of variances metric to estimate disparity here.
Note that there are still statistical issues with this metric, but for the purposes of comparison with previous work we want to use a standard metric for these analyses.

\paragraph{Time subsampling} 
\label{time_subsamples}

To estimate disparity-through-time we first need to split the data into time subsamples.
Here we explore the following scenarios.

\begin{enumerate}
  \item Stratigraphic time bins. This is the traditional method, where all the taxa within each stratigraphic period are included in the disparity calculation. This often leads to bins of unequal duration. Here we use stratigraphic ages and epochs.
  \item Equally sized time bins. This is another commonly used method, where the time frame of interest is split into equally sized time bins, then all the taxa within each time bin are included in the disparity calculation. Here we use three different methods to determine the size of the time bins. 
    \begin{enumerate}
      \item The \textit{duration} of the bin is equal to the median duration of the stratigraphic period (age or epoch).
      \item The \textit{number} of bins is equal to the number of stratigraphic periods (ages or epochs).
      \item The \textit{number} of bins is either 2, 5, 10, 15, or 20.
    \end{enumerate}
  \item Time slicing. We describe this in more detail below, but in brief, time slicing uses a phylogeny, and rather than binning the data, it takes slices through a phylogeny and includes all the taxa and nodes in that slice within the disparity calculation. Here we use three different methods to determine the intervals between the time slices.
    \begin{enumerate}
      \item The \textit{interval} between slices is equal to the median duration of the stratigraphic period (age or epoch).
      \item The \textit{number} of slices is equal to the number of stratigraphic periods (ages or epochs).
      \item The \textit{number} of slices is either 2, 5, 10, 15, or 20.
    \end{enumerate}

\paragraph{Time slicing} 
\label{time_slicing}
The ``time-slicing'' approach considers subsets of taxa in the cladisto-space at specific equidistant points in time, as opposed to considering subsets of taxa between two points in time.
This results in even-sampling of the cladisto-space across time and while permitting to use an underlying model of character evolution (punctuated or gradual).

In practice, time-slicing considers the disparity of any element present in the phylogeny (branches, nodes and tips) at any point in time.
When the phylogenetic elements are nodes or tips, the ordination scores for the nodes (estimated using ancestral state reconstruction as described above) or tips are directly used for estimating disparity.
When the phylogenetic elements are branches we choose the ordination score for the branch using one of two evolutionary models:

@@

    \item{\textbf{Punctuated evolution.}} 
    This model selects the ordination score from either the ancestral node or the descendant node/tip of the branch regardless of the position of the slice along the branch. 
    Similarly to the time bin approach, this reflects a model of punctuated evolution where changes in disparity occur either at the start or at the end of a branch over a relatively short time period and clades undergo long periods of stasis during their evolution \citep{Gould1977,Hunt20112007}.
    We apply this model in four ways:

@@

      \item The ``acctran'' model, always selecting the ordination score of the ancestral node of the branch.
      \item The ``deltran'' model, always selecting the ordination score of the descendant node/tip of the branch.
      \item The ``random'' model, randomly selecting the ordination score of either the ancestor or the descendant of the branch.
      \item The ``proximity'' model, selecting the ordination of either the ancestor or the descendant if the slice occurs respectively before or after half the branch length.

@@

    The two first models assume that changes always occur early (ACCelerated TRANsition) or late along the branches (DELayed TRANsition).
    The third model makes neither assumption and simply selects randomly the data between the ancestor and the descendant and the fourth base the selection of either the ancestor or the descendant on where the slice occurs along the branch.
    The punctuated models only select either the ordination score from the ancestor and the descendant once in the whole disparity analysis.
    For example, if using the ``random'' model, if the data of the ancestor has been randomly chosen, only this data will be used during the bootstraps procedure and for the disparity calculation.

    \item{\textbf{Gradual evolution.}}
    Contrary to the punctuated models, these following models do not select the ordination score of either the ancestor or the descendant but associate a probability to both.
    This reflects a model of gradual evolution where changes in disparity are gradual and cumulative along the branch.

    \begin{enumerate}[(i)]
    \setcounter{enumii}{4}
      \item The ``punctuated'' model (probabilistic), selecting the ordination score from both the ancestor and the descendant with an equal probability
          \begin{equation}
          p(\text{ancestor}) = p(\text{descendant}) = 0.5
          \end{equation}

    \item The ``gradual'' model (probabilistic), selecting the ordination score from both the ancestor and the descendant with a probability equal to the length between the slice.
          \begin{equation}
              p(\text{ancestor}) = \frac{d(\text{ancestor},\text{slice})}{d(\text{ancestor},\text{descendant})}
          \end{equation}
          \begin{equation}
              p(\text{descendant}) = 1 - p(\text{ancestor})
          \end{equation}
    \noindent Where $d(x,y)$ is the distance measured as branch length.
    \end{enumerate}

    In these models, both the ancestor's and descendant's data will participate to the disparity calculation based on the following models:
    For example, using the ``gradual.split'' model, if slice occurs at the third quarter of a branch from A to B (75\% of the total branch length), after bootstrapping, the disparity results will be based on 25\% of the data from A and 75\% of the data from B.
    Because of the probabilistic nature of these models, there are only meaningful when calculating disparity from bootstrapped datasets.
\end{enumerate}

%TG: Justification on the fact that these models are not ACEs.
It is important to note that the time-slicing method is not an ancestral states estimation method \textit{per se}.
This method does not estimate values along a branch applying a model (\textit{c.f.} methods described in \ref{ace}) but rather chooses between the two available pieces of information (the ordination score of the descendant or the ancestor) using the models described above.
This allows this method to be used in post-ordination analysis were the data used in each time slice is data already present in the cladisto-space.
In other words, this method does not require a re-ordination of the cladisto-space every time a slice goes through a branch, thus allowing the properties of the cladisto-space (e.g. distance between species, variance per axis, etc.) to remain constant.

@@

\subsection{Estimating disparity-through-time: bootstrapping}
Finally, to reduce the influence of outliers on our disparity estimates, we bootstrapped each disparity measurement for each time subsampling method, by randomly resampling with replacement a new subsample of taxa from the observed taxa in the subsample 1000 times.
We then calculated the median disparity value for each subsample along with the 50\% and 95\% confidence intervals.
We also recorded the number of taxa (or taxa and nodes for time slicing methods) in each subsample as a proxy for taxonomic diversity (Table \ref{tab:subsamples}).

@@

\section{Acknowledgments}
	We thank all the PalAss folk

>>>>

To address these issues, we propose a ``time-slicing'' approach that takes advantage of the wealth of paleontological datasets which now have associated phylogenies. 
Time-slicing uses a phylogenetic tree and considers subsets of taxa at specific equidistant points in time, as opposed to considering subsets of taxa between two points in time.
This results in even-sampling across time and permits us to define the underlying model of character evolution (punctuated or gradual).  
Time-slicing also includes any element present in the phylogeny (branches, nodes and tips) at the time slice in question as part of the disparity calculation.
This allows us to measure disparity at time points where there are no terminal taxa, and increases the sample size at each time point, making downstream analyses of the drivers of disparity much more feasible.

Here we present our time-slicing methods using four datasets taken from the literature.
We calculate disparity-through-time for each dataset using a range of time binning and time-slicing methods, and then compare these approaches with respect to the relative disparities calculated, but also with how the different approaches influence biological conclusions. We find that XXX

\section{Materials and Methods}
\subsection{Overview}
\label{overview-section}
To test the different time subsampling methods, we followed the protocol below (Figure \ref{datasets}). 
All the code needed to reproduce these analyses (along with detailed instructions) is provided on GitHub (https://github.com/nhcooper123/time-slice). 
% NC: might move this to the NHM github account

% Add overview figure

  %\begin{figure}[!htbp]
    %\centering
    %\includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/overview}
    %\caption[caption title]
    % {caption text}
    %\label{overview}
  %\end{figure}

\subsection{Example datasets}
\label{datasets}
To test the different time binning/slicing methods we selected four datasets: a mammal dataset from \cite{beckancient2014}, two theropod datasets from \cite{brusatte2014gradual} and \cite{bapst2016topology}, and a crinoid dataset from \cite{wright2017bayesian}.
Table \ref{datasets} and appendix ??? provide more details. % add appendix details
Each dataset consists of first and last occurrence dates for all taxa, a matrix of morphological characters in NEXUS format, and a time-scaled phylogeny. 
These datasets are freely available with their accompanying papers (Table \ref{datasets}), but for reproducibility purposes we also provide the data we used on the NHM Data Portal (CITE).

% Dataset details table
  \begin{table}[!htbp]      
    \caption[Datasets.]{caption}
    \input{tables/table_datasets.tex} 
    \label{datasets}  
  \end{table}

\subsection{Preparing the data for disparity-through-time analysis}

\paragraph{Estimating ancestral character states.}
\label{ace}
For each dataset we used the re-rooting method \citep{Yang01121995,Garland2000} to get Maximum Likelihood estimates of the ancestral states for each character at every node in the phylogeny, using the \texttt{rerootingMethod} function from the \texttt{R} package \texttt{phytools} version 0.6-44 \citep{phytools,R}.
Where there were missing character data for a taxon we followed the method of \cite{Claddis} and treated missing data as any possible observed state for each character.
For example, if a character had two observed states ($0$ and $1$) across all taxa, we attributed the multi-state ``$0$\&$1$" value to the taxon with missing data, representing an equal probability of being either $0$ or $1$.
This allows the ancestral node of a taxon with missing data to be estimated with no assumptions other than that the taxon has one of the observed character states.
To prevent poor ancestral state reconstructions from biasing our results, especially when a lot of error is associated with the reconstruction, we only included ancestral state reconstructions with a scaled Likelihood $\geq$ $0.95$.
Ancestral state reconstructions with scaled Likelihoods below this threshold were recoded as missing data (``?'').

@@

\paragraph{Constructing distance matrices.}
To estimate the cladisto-spaces for each of our datasets we first constructed pairwise distance matrices of length $k$, where $k$ is the total number of tips and nodes in the dataset.
We calculated the $k$$\times$$k$ distances using the Gower distance \citep{Gower71}, i.e. the Euclidean distance between two taxa divided by the number of shared characters. 
This allows us to correct for distances between two taxa that share many characters and could be closer to each other than to taxa with fewer characters in common (i.e. because some pairs of taxa share more characters in common than others, they are more likely to be similar).
For cladistic matrices, using this corrected distance is preferable to the raw Euclidean distance because of its ability to deal with discrete or/and ordinated characters as well as with missing data \citep{anderson2012using}.
However, the Gower distance cannot calculate distances when taxa have no overlapping data.
Therefore, we used the \texttt{TrimMorphDistMatrix} function from the \texttt{Claddis} \texttt{R} package \citep{Claddis} to remove pairs of taxa with no cladistic characters in common.
This led to us removing nine taxa from the \cite{bapst2016topology} dataset, and 19 from the \cite{brusatte2014gradual} dataset, but none from the other two datasets (see appendix for details of which species). 
% add appendix details

@@

\subsection{Disparity-through-time analyses}

\paragraph{Calculating disparity.}
\label{disparity_calc}
Disparity can be calculated in many different ways \citep[e.g.][]{Wills1994,Ciampaglio2004,thorneresetting2011,hopkinsdecoupling2013,huang2015origins}, however most studies estimate disparity using four metrics: the sum and products of ranges and variances, each of which gives a slightly different estimate of how the data fits within the cladisto-space \citep{Foote01071994,Wills1994,brusatte50,Brusatte12092008,cisneros2010,thorneresetting2011,prentice2011,brusattedinosaur2012,toljagictriassic-jurassic2013,ruta2013,bentonmodels2014,bensonfaunal2014}.
However, these metrics have limitations. 
First, the range metrics are affected by the uneven sampling of the fossil record \citep{Butler2012}.
Second, because we include all $n$ dimensions in the analysis (see above), the products of ranges and variances will tend towards zero since the scores of the last dimension are usually really close to zero themselves. 
We therefore use the sum of variances metric to estimate disparity here.
Note that there are still statistical issues with this metric, but for the purposes of comparison with previous work we want to use a standard metric for these analyses.

\paragraph{Time sub-sampling} 
\label{time_subsamples}

To estimate disparity-through-time we first need to split the data into time subsamples.
Here we use three time sub-sampling methods.

\begin{enumerate}
  \item Stratigraphic time bins. This is the traditional method, where all the taxa within each stratigraphic period are included in the disparity calculation. This often leads to bins of unequal duration. Here we use stratigraphic ages and epochs.
  \item Equally sized time bins. This is another commonly used method, where the time frame of interest is split into equally sized time bins, then all the taxa within each time bin are included in the disparity calculation. 
  \item Time-slicing. We describe this in more detail below, but in brief, time-slicing uses a phylogeny, and rather than binning the data, it takes slices through a phylogeny and includes all the taxa and nodes in that slice within the disparity calculation. 
\end{enumerate}

\paragraph{Time-slicing} 
\label{time_slicing}
The ``time-slicing'' approach considers subsets of taxa in the cladisto-space at specific equidistant points in time, as opposed to considering subsets of taxa between two points in time.
This results in even-sampling of the cladisto-space across time and allows us to use different underlying models of character evolution (punctuated or gradual).

In practice, time-slicing considers the disparity of any element present in the phylogeny (branches, nodes and tips) at any point in time.
When the phylogenetic elements are nodes or tips, the ordination scores for the nodes (estimated using ancestral state reconstruction as described above) or tips are directly used for calculating disparity.
When the phylogenetic elements are branches we choose the ordination score for the branch using one of two evolutionary models:

@@

    \item{\textbf{Punctuated evolution.}} 
    This model selects the ordination score from either the ancestral node or the descendant node/tip of the branch regardless of the position of the slice along the branch. 
    Similarly to the time bin approach, this reflects a model of punctuated evolution where changes in disparity occur either at the start or at the end of a branch over a relatively short time period, and clades undergo long periods of stasis during their evolution \citep{Gould1977,Hunt20112007}.
    We apply this model in four ways:

@@

      \item The ``acctran'' model, always selecting the ordination score of the ancestral node of the branch.
      \item The ``deltran'' model, always selecting the ordination score of the descendant node/tip of the branch.
      \item The ``random'' model, randomly selecting the ordination score of either the ancestor or the descendant of the branch.
      \item The ``proximity'' model, selecting the ordination score of the ancestor if the slice occurs in the first half of the branch, and the descendant if the slice occurs in the second half of the branch.

@@

    The two first models assume that changes always occur early (ACCelerated TRANsition) or late along the branches (DELayed TRANsition).
    The third model makes neither assumption and simply selects data from the ancestor or the descendant at random, and the fourth bases the selection of either the ancestor or the descendant on where the slice occurs along the branch.
    The punctuated models only select either the ordination score from the ancestor and the descendant once in the whole disparity analysis.
    For example, if using the ``random'' model, if the data of the ancestor has been randomly chosen, only this data will be used during the bootstrapping (see below) and for the disparity calculation.

    \item{\textbf{Gradual evolution.}}
    Unlike the punctuated models, the following models do not select the ordination score of either the ancestor or the descendant but associate a probability to both.
    This reflects a model of gradual evolution where changes in disparity are gradual and cumulative along the branch.

    \begin{enumerate}[(i)]
    \setcounter{enumii}{4}
      \item The ``equal splits'' model (probabilistic), selects the ordination score from both the ancestor and the descendant with an equal probability
          \begin{equation}
          p(\text{ancestor}) = p(\text{descendant}) = 0.5
          \end{equation}

    \item The ``gradual splits'' model (probabilistic), selects the ordination score from both the ancestor and the descendant with a probability equal to the distance between the slice.
    % NC: I'm unsure what you mean here: "the distance between the slice." Do you mean the branch length, the interval between slices, or the distance between the slice and the node?
          \begin{equation}
              p(\text{ancestor}) = \frac{d(\text{ancestor},\text{slice})}{d(\text{ancestor},\text{descendant})}
          \end{equation}
          \begin{equation}
              p(\text{descendant}) = 1 - p(\text{ancestor})
          \end{equation}
    \noindent Where $d(x,y)$ is the distance measured as branch length.
    \end{enumerate}

    In these models, both the ordination scores of both the ancestor and descendant contribute to the disparity calculation based on the following models:
    For example, using the ``gradual.split'' model, if slice occurs at the third quarter of a branch from A to B (75\% of the total branch length), after bootstrapping, the disparity results will be based on 25\% of the data from A and 75\% of the data from B.
    Because of the probabilistic nature of these models, there are only meaningful when calculating disparity from bootstrapped datasets.
\end{enumerate}

%TG: Justification on the fact that these models are not ACEs.
It is important to note that the time-slicing method is not an ancestral states estimation method \textit{per se}.
This method does not estimate values along a branch applying a model (\textit{c.f.} methods described for reconstructing ancestral states in the preparing data for disparity-through-time section above) but rather chooses between the two available pieces of information (the ordination score of the descendant or the ancestor) using the models described above.
This allows the method to be used in post-ordination analysis where the data used in each time slice is data already present in the cladisto-space.
In other words, this method does not require a re-ordination of the cladisto-space every time a slice goes through a branch, thus allowing the properties of the cladisto-space (e.g. distance between species, variance per axis, etc.) to remain constant.

\begin{enumerate}
  \item Stratigraphic time bins. This is the traditional method, where all the taxa within each stratigraphic period are included in the disparity calculation. This often leads to bins of unequal duration. Here we use stratigraphic ages and epochs.
  \item Equally sized time bins. This is another commonly used method, where the time frame of interest is split into equally sized time bins, then all the taxa within each time bin are included in the disparity calculation. Here we use three different methods to determine the size of the time bins. 
    \begin{enumerate}
      \item The \textit{duration} of the bin is equal to the median duration of the stratigraphic period (age or epoch).
      \item The \textit{number} of bins is equal to the number of stratigraphic periods (ages or epochs).
      \item The \textit{number} of bins is either 2, 5, 10, 15, or 20.
    \end{enumerate}
  \item Time-slicing. We describe this in more detail below, but in brief, time-slicing uses a phylogeny, and rather than binning the data, it takes slices through a phylogeny and includes all the taxa and nodes in that slice within the disparity calculation. Here we use three different methods to determine the intervals between the time slices.
    \begin{enumerate}
      \item The \textit{interval} between slices is equal to the median duration of the stratigraphic period (age or epoch).
      \item The \textit{number} of slices is equal to the number of stratigraphic periods (ages or epochs).
      \item The \textit{number} of slices is either 2, 5, 10, 15, or 20.
    \end{enumerate}

Applying the models

To compare our three approaches (stratigraphic time bins, equal time bins and time-slicing) we applied the methods as follows.

\begin{enumerate}
  \item Stratigraphic time bins. We calculated disparity for each stratigraphic 
\begin{enumerate}
  \item Unequal time bins and unevenly spaced time slices, where the bins are stratigraphic periods (ages or epochs) and slices occur at the midpoint of the stratigraphic period.
  \item Equal time bins and equally spaced time slices, where time bin size, or the interval between slices, is defined by the average \textit{duration} of the stratigraphic period (age or epoch).
  \item Equal time bins and equally spaced time slices, where the number of time bins, or number of slices, is defined by the average \textit{number} of stratigraphic periods (ages or epochs) in the time frame of interest.
\end{enumerate}

  \item Equally sized time bins. This is another commonly used method, where the time frame of interest is split into equally sized time bins, then all the taxa within each time bin are included in the disparity calculation. Here we use three different methods to determine the size of the time bins. 
    \begin{enumerate}
      \item The \textit{duration} of the bin is equal to the median duration of the stratigraphic period (age or epoch).
      \item The \textit{number} of bins is equal to the number of stratigraphic periods (ages or epochs).
      \item The \textit{number} of bins is either 2, 5, 10, 15, or 20.
    \end{enumerate}
  \item Time-slicing. We describe this in more detail below, but in brief, time-slicing uses a phylogeny, and rather than binning the data, it takes slices through a phylogeny and includes all the taxa and nodes in that slice within the disparity calculation. Here we use three different methods to determine the intervals between the time slices.
    \begin{enumerate}
      \item The \textit{interval} between slices is equal to the median duration of the stratigraphic period (age or epoch).
      \item The \textit{number} of slices is equal to the number of stratigraphic periods (ages or epochs).
      \item The \textit{number} of slices is either 2, 5, 10, 15, or 20.
    \end{enumerate}

@@

\subsection{Estimating disparity-through-time: bootstrapping}
Finally, to reduce the influence of outliers on our disparity estimates, we bootstrapped each disparity measurement for each time subsampling method, by randomly resampling with replacement a new subsample of taxa from the observed taxa in the subsample 1000 times.
We then calculated the median disparity value for each subsample along with the 50\% and 95\% confidence intervals.
We also recorded the number of taxa (or taxa and nodes for time-slicing methods) in each subsample as a proxy for taxonomic diversity (Table \ref{tab:subsamples}).

@@

\section{Acknowledgments}
	We thank all the PalAss folk.
  TG acknowledge support from the Australian Discovery Project Grant number DP170103227 awarded to Vera Weisbecker.
